\begin{abstract}
Probabilistic programming languages (PPLs) are powerful tools for statistical modeling, but performing exact inference on programs with continuous random variables remains a major challenge. Many PPLs capable of exact inference are restricted to purely discrete models, while those supporting continuous distributions typically rely on approximate inference. This paper introduces \Slice{}, a language and compiler that bridges this gap by automatically and soundly transforming probabilistic programs with continuous distributions into equivalent, purely discrete counterparts. The core of our approach is a novel type system that statically analyzes how continuous variables are compared against constant values throughout the program. This type-directed analysis determines a set of cut points used to partition continuous domains into a finite number of intervals. \Slice{} then replaces the original continuous distributions with discrete distributions over these intervals, enabling the transformed program to be solved by off-the-shelf exact inference engines for discrete PPLs. However, this approach has a fundamental limitation: it can only fully discretize programs where continuous variables are compared to constants. Programs comparing continuous variables to each other can only be partially discretized or remain continuous, limiting the applicability of exact discrete inference. We formalize this transformation, prove its soundness, and provide an open-source implementation. Our evaluation on a range of benchmarks demonstrates that for fully discretizable programs, this approach is significantly faster than existing symbolic inference techniques for hybrid programs while maintaining correctness.
\end{abstract}